---
title: "What does the data tell us about Toronto bike theft risks and prevention?"
author: "Talia Fabregas"
format: html
---

```{r, echo=FALSE, message=FALSE, warning=FALSE, include=FALSE}
library(dplyr)

# Load the data
knitr::opts_chunk$set(echo=FALSE, include=TRUE, eval=TRUE)
required_packages <- c(
  "lubridate", "sf", "tidyverse"
  # include any packages to be installed here
) 
for (p in required_packages) { # install if missing
  if (!require(p, character.only = TRUE)) {
    install.packages(p, character.only = TRUE)
  }
}

library(tidyverse)
library(sf)
library(lubridate)
library(ggplot2)
library(plotly)

# bike thefts csv file
bike_thefts_clean <- read_csv("cleaned_data/bike_thefts_clean.csv")

# load the bike thefts GeoJSON file
# bike_thefts_sf <- read_sf("cleaned_data/bike_thefts_clean_sf.geojson")

# bike parking geojson file
bike_parking_sf <- read_sf("data/bike_parking_map_data.geojson")

# toronto neighbourhoods map data
toronto_neighbourhoods_shp <- st_read("data/Neighbourhoods - 4326/Neighbourhoods - 4326.shp")

# police divisions map data
police_divisions_shp <- st_read("data/PoliceBoundariesData-4326/Police Boundaries Data - 4326.shp")

# GeoJSON format maybe useful if you would like to combine
# an external geospatial dataset
bike_thefts_sf <- read_sf("data/Bicycle_Thefts_Open_Data.geojson") |>
  mutate(
    occurence_date = as_date(
      OCC_DATE, format = "%a, %d %b %Y %H:%M:%S GMT"
    ),
    report_date = as_date(
      REPORT_DATE, format = "%a, %d %b %Y %H:%M:%S GMT"
    )
  ) |>
    filter(REPORT_YEAR >= 2014 & 
           REPORT_YEAR <= 2023 & 
           !is.na(NEIGHBOURHOOD_158) &
           !is.na(STATUS) & 
           !is.na(PREMISES_TYPE)
  ) 

bike_parking_sf <- st_read("data/bike_parking_map_data.geojson")
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
year_month_summary <- bike_thefts_clean %>%
  group_by(REPORT_YEAR, OCC_MONTH) %>%
  summarise(Total_Thefts = n_distinct(OBJECTID), .groups = "drop")

report_summary <- year_month_summary %>%
  group_by(REPORT_YEAR) %>%
  summarise(YEARLY_TOTAL = sum(Total_Thefts))

neighbourhood_counts <- bike_thefts_sf |>
  group_by(NEIGHBOURHOOD_158) |>
  summarise(num_thefts = n_distinct(OBJECTID),
            lat = mean(LAT_WGS84),
            long = mean(LONG_WGS84),
            .groups="drop")

division_counts <- bike_thefts_sf |>
  group_by(DIVISION) |>
  summarise(num_thefts = n_distinct(OBJECTID),
            .groups="drop")

thefts_summary <- bike_thefts_clean %>%
  group_by(REPORT_YEAR, PREMISES_TYPE) %>%
  summarise(count = n(), .groups = "drop") %>%
  mutate(proportion = count / sum(count)) %>%
  mutate(percentage = (count / sum(count)) * 100) %>%
  ungroup()

thefts_per_year <- thefts_summary %>%
  group_by(REPORT_YEAR) %>%
  summarise(total_thefts = sum(count), .groups = "drop")

top_10_neighbourhoods <- neighbourhood_counts %>%
  arrange(desc(num_thefts)) %>%
  slice_head(n = 10)

recovered_bikes <- bike_thefts_clean |>
  mutate(is_recovered = ifelse(STATUS == "RECOVERED", 1, 0)) 

# you're less likely to get a stolen bike back now than you were 10 years ago
recovered_bikes_yearly <- recovered_bikes |>
  group_by(REPORT_YEAR) |>
  summarise(
            num_thefts = n_distinct(OBJECTID),
            num_not_recovered = n_distinct(OBJECTID) - sum(is_recovered),
            num_recovered = sum(is_recovered),
            percent_recovered  = round((mean(is_recovered) * 100), 2),
            prop_recovered = mean(is_recovered),
            .groups="drop")

# no bikes in 8 of the 10 top neighbourhoods were recovered in 2023
recovered_bikes_top10 <- recovered_bikes |>
  filter(NEIGHBOURHOOD_158 %in% top_10_neighbourhoods$NEIGHBOURHOOD_158) |>
  group_by(NEIGHBOURHOOD_158, REPORT_YEAR) |>
  summarise(
    num_thefts = n_distinct(OBJECTID),
    num_recovered = sum(is_recovered),
    num_not_recovered = n_distinct(OBJECTID) - sum(is_recovered),
    prop_recovered = mean(is_recovered),
    pct_recovered = round((mean(is_recovered) * 100), 2), 
            .groups="drop")
```

Toronto commuters know traffic and public transit delays all too well. Cycling can be an efficient, healthy, and convenient way to get around the city. It can be faster than walking, sitting in traffic, or dealing with transit delays and it’s an excellent aerobic exercise for adults different ages and skill levels. As a Toronto cyclist myself, I know that many of my fellow cyclists are worried about their bikes being stolen. We hear about bike thefts all the time and we’re all familiar with the narrative that bikes are more likely to be stolen downtown. Is your bike really more likely to get stolen in downtown, or is the higher number of bike thefts downtown simply due to higher cyclist volume?

# A snapshot of bike theft trends from the past decade

The good news is that bike thefts are not on the rise in the City of Toronto. The bad news is that if your bike gets stolen, you’re probably never going to get it back. No matter where in Toronto it was stolen.

```{r}
#| echo: false
#| message: false
#| warning: false
#| fig-subcap: ["Toronto bike thefts peaked in 2018, but have been trending downwards since 2020. There were fewer bike thefts in 2022 and 2023 than in 2014", "Once a bike is stolen, it is probably gone forever. Stolen bikes are even less likely to be found now than in 2014."]
#| layout-ncol: 2
# Plot the line graph showing the total number of thefts per year
p1 <- ggplot(thefts_per_year, aes(x = REPORT_YEAR, y = total_thefts)) +
  geom_line(color = "#980043", size = 1) +  # Line graph
  geom_point(color = "#980043", size = 2) +   # Points for each year
  geom_hline(yintercept = 3499, linetype = "dashed", color = "#333335", size = 1) + 
    annotate("text", x = 2021, y = 3650, label = "Average: 3499", color = "#333335", size = 4, hjust = 0) + # Annotation
  labs(
    title = "How many bike thefts have there been every year?",
    x = "Year",
    y = "Total # Bike Thefts"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16),
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  scale_y_continuous(limits = c(2500, 4500)) +
  scale_x_continuous(breaks = seq(2014, 2023, by = 1)) 

p2 <- ggplot(recovered_bikes_yearly, aes(x = REPORT_YEAR, y = prop_recovered)) +
  geom_line(color = "#80b1d3", size = 1) +  # Line style and color
  geom_point(color = "#80b1d3", size = 2.5) +   # Points for each year
  labs(
    title = "Less than 1% of Bikes Stolen Since 2021 Were Recovered",
    x = "Year",
    y = "% of stolen bikes recovered"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
    axis.text.y = element_text(size = 10)
  ) +
  scale_y_continuous(labels = scales::percent_format(), limits = c(0, 0.025)) +
  scale_x_continuous(breaks = seq(2014, 2023, by = 1))  # Ensure all years appear

print(p1)
print(p2)
```

Since 2014, an average of 3499 bikes were stolen each year in Toronto. My analysis found that bike thieves are almost always successful, but they have not been emboldened over the last decade. Less than 1% of bikes stolen since 2021 were recovered and returned to the owner. From 2014 until the peak in 2018, bike thefts were increasing in Toronto. However, bike thefts decreased between 2020 and 2022, and there were fewer bike thefts in 2023 than in 2014.

Even if bike thefts are not on the rise, knowing where they happen is an important step towards protecting your bike. Hover over the figure below to see the percentage of bike thefts that occurred at apartments/houses, outside, commercial, educational , transit, and other premises.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-premiseinteractive
#| fig-cap: Most bikes are stolen from a residence (apartment or house) or outside, but the percentage of Toronto bike thefts each year per premise type has not changed much 2014. . 


library(plotly)

thefts_summary <- thefts_summary %>%
  group_by(REPORT_YEAR) %>%
  mutate(total_thefts = sum(count),  # Total number of thefts for the year
         percentage = (count / total_thefts)) %>%  # Calculate percentage
  ungroup() %>%  # Remove grouping
  mutate(REPORT_YEAR = factor(REPORT_YEAR, levels = sort(unique(REPORT_YEAR))))

custom_colors2 <- rev(c("#8dd3c7", "#ffffb3", "#bebada", "#fb8072", "#80b1d3", "#fdb462", "#b3de69"))

thefts_summary$PREMISES_TYPE <- factor(thefts_summary$PREMISES_TYPE,
                                       levels = c("Apartment", "House", "Outside", "Commercial", "Educational", "Transit", "Other"))

# Plot using plotly with the custom colors
plot_ly(thefts_summary, x = ~REPORT_YEAR, y = ~percentage, color = ~PREMISES_TYPE,
        colors = custom_colors2,  # Apply the custom color array here
        type = 'bar',
        text = ~paste(PREMISES_TYPE, ": ", scales::percent(percentage, accuracy = 0.01), sep = ""),
        hoverinfo = 'text',
        scalegroup = ~REPORT_YEAR) %>%
  layout(
    title = "Where are bikes most commonly stolen in Toronto?",
    xaxis = list(title = "Year"),
    yaxis = list(title = "Percent of Toronto Bike Thefts", tickformat = ".0%"),
    barmode = 'stack',
    legend = list(title = list(text = "Premise Type")),
    colorway = custom_colors2  # Make sure this is included to apply the colorway
  ) 

```

This graph reveals some important insights about bike thieves' favorite premises. Bike thieves love to target bikes left unattended at cyclists' residences (both houses and apartments) and outside, but they have not had much success on commercial, educational, or transit premises. Commercial, educational, and transit premises often have staff and security cameras. Most houses, apartments, and outdoor areas where bikes can be parked do not. The majority of bike thefts reported since 2015 have taken place at a house, apartment, or outside. In 2023, nearly twelve times more bikes were stolen from residences (apartments and houses combined) than from transit premises.

Since 2015, **more bike thefts have taken place at a residence** (house or apartment) **than any other premise**. In 2014, nearly one-third more bike thefts took place at houses than apartments. In 2024, there were nearly twice as many bike thefts at apartments compared to houses. The overall share of bike thefts that took place residences (including both houses and apartments) remained stable between 2014 and 2023 but it peaked in 2020 when many people stayed home due to the Covid-19 pandemic. The share of residence bike thefts at apartments has increased over time, while the share of residence bike thefts at houses has decreased over time. According to 2017 census data, more Torontonians live in apartments than houses, but it is difficult to draw a conclusion about risk of bike theft based on type of residence because the data does not reveal anything the number of Toronto cyclists who live in apartments versus houses. There is no evidence that living in a house or apartment puts a cyclist in Toronto at a higher risk of having their bike stolen, but one thing is clear: **more bikes are stolen from cyclists' homes than any other premise.** Securing your bike at home is an easy and effective way to prevent it from being stolen, regardless of the type of home you live in.

# What you really need to know about downtown bike thefts

Yes, the highest concentration of bike thefts in 2023 is in and around the downtown core. No, this does not mean your bike is more likely to get stolen downtown.

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-interactivemap
#| fig-cap: If we use high-capacity bike parking locations to measure cyclist volume, the Toronto neighbourhoods with the most bike thefts in 2023 are the ones with the highest cyclist volume. 
library(plotly)

bike_thefts_2023 <- bike_thefts_sf |>
  filter(REPORT_YEAR == 2023 & LONG_WGS84 < -79 & LAT_WGS84 > 40)
  # filter(OCC_YEAR >= 2014 & OCC_YEAR <= 2023 & LONG_WGS84 < -79 & LAT_WGS84 > 40)
  

bike_sf <- st_as_sf(bike_thefts_2023, coords = c("LONG_WGS84", "LAT_WGS84"), crs = 4326)

bike_sf$NEIGHBOURHOOD_158 <- gsub("\\s*\\(.*\\)", "", bike_sf$NEIGHBOURHOOD_158)


toronto_neighbourhoods_shp2 <- toronto_neighbourhoods_shp 

toronto_neighbourhoods_shp2$AREA_DE8 <- gsub("\\s*\\(.*\\)", "", toronto_neighbourhoods_shp$AREA_DE8)

theft_map <- ggplot()+
    geom_sf(data = toronto_neighbourhoods_shp2, 
            aes(text = paste(AREA_DE8)),
            fill = "lightgray", 
            color = "black", 
            size = 0.3, alpha=0.3)  +
    geom_sf(data = bike_sf, 
            aes(text = paste(NEIGHBOURHOOD_158)),
            color = "#980043", 
            size = 0.7, 
            alpha = 0.2) +
    theme_minimal() +
 theme_minimal() +
    theme(
      panel.grid = element_blank(),  # Remove grid lines
      panel.background = element_blank(),
      axis.text = element_blank(),  # Remove axis text
      axis.ticks = element_blank(),  # Remove axis ticks
      axis.title = element_blank(),
      legend.title = element_text(size = 12),  # Customize legend title size
      legend.text = element_text(size = 10),   # Customize legend text size
      legend.position = "bottom"  # Position legend at the bottom
    ) +
    ggtitle("Most Toronto bike thefts occur in busiest neighbourhoods")

theft_map <- ggplotly(theft_map, tooltip="text") %>%
     layout(
      hoverlabel = list(
      bgcolor = "lightgray",  # White background
      font = list(color = "black")  # Black text for contrast
      ),
      legend = list(
      orientation = "h",       # Horizontal legend
      x = 0.7,                 # Centered horizontally
      y = 0.2,                # Place it under the map
      xanchor = "center",      # Align to center
      yanchor = "top"          # Align legend's top with the position
    ),
    margin = list(t = 50, r = 20, b = 100, l = 20)
    )

bike_parking_capacity <- bike_parking_sf |> 
  filter(!is.na(BICYCLE_CAPACITY) & BICYCLE_CAPACITY > 0)

bike_parking_capacity$BICYCLE_CAPACITY <- as.numeric(bike_parking_capacity$BICYCLE_CAPACITY)


# Remove legends from both maps
parking_map <- ggplot() +
    # Add neighborhood polygons
    geom_sf(data = toronto_neighbourhoods_shp2,
            aes(text = paste(AREA_DE8)),
            fill = "lightgray", 
            color = "black", 
            size = 0.3, alpha = 0.3) +
    # Add bike parking locations with no legend
    geom_sf(data = bike_parking_capacity, 
            aes(size = sqrt(BICYCLE_CAPACITY), color = "Parking Spot"), 
            alpha = 0.4) +
    # Set a minimal theme
    theme_minimal() +
    theme(
      panel.grid = element_blank(),  # Remove grid lines
      panel.background = element_blank(),
      axis.text = element_blank(),  # Remove axis text
      axis.ticks = element_blank(),  # Remove axis ticks
      axis.title = element_blank(),
      legend.position = "none"  # Remove legend
    ) +
    # Define scales for color and size (without legend)
    scale_color_manual(
      name = "", 
      values = c("Parking Spot" = "steelblue")
    ) +
    scale_size_continuous(
      name = "",  # Legend title
      range = c(1, 1.5),  # Adjust the point size range as needed
      breaks = c(5, 10, 20, 40, 60, 80, 100, 120),  # Adjust breaks to match your data range
      labels = c("5", "10", "20", "40", "60", "80", "100", "120")
    ) +
    ggtitle("High-Theft Neighbourhoods Align with Top Bike Parking Spots")

# Convert to plotly for interactivity
parking_map <- ggplotly(parking_map, tooltip = "text") %>%
    layout(
      hoverlabel = list(
        bgcolor = "lightgray",  
        font = list(color = "black")
      ),
      margin = list(t = 50, r = 20, b = 100, l = 20)  # Add space for the caption below
    )

# Remove legends from the theft map
theft_map <- ggplot() +
    geom_sf(data = toronto_neighbourhoods_shp2, 
            aes(text = paste(AREA_DE8)),
            fill = "lightgray", 
            color = "black", 
            size = 0.3, alpha=0.3)  +
    geom_sf(data = bike_sf, 
            aes(text = paste(NEIGHBOURHOOD_158), color = "Bike Theft Location"), 
            size = 0.5, 
            alpha = 0.2) +
    theme_minimal() +
    theme(
      panel.grid = element_blank(),  # Remove grid lines
      panel.background = element_blank(),
      axis.text = element_blank(),  # Remove axis text
      axis.ticks = element_blank(),  # Remove axis ticks
      axis.title = element_blank(),
      legend.position = "none"  # Remove legend
    ) +
    scale_color_manual(
      name = "", 
      values = c("Bike Theft Location" = "#980043")  # Set color for bike theft locations
    ) +
    ggtitle("Where in Toronto did bike thefts happen in 2023?")

theft_map <- ggplotly(theft_map, tooltip = "text") %>%
    layout(
      hoverlabel = list(
        bgcolor = "lightgray",  # White background
        font = list(color = "black")  # Black text for contrast
      )
    )

# Combine both maps with captions
combined_plot <- subplot(
    theft_map %>% layout(
        annotations = list(
            x = 0.2, y = 0,
            text = "Locations of Bike Thefts Reported in 2023",  # Caption for theft_map
            showarrow = FALSE,
            xref = "paper", yref = "paper",
            font = list(size = 10, color = "black"),
            align = "center"
        )
    ),
    parking_map %>% layout(
        annotations = list(
            x = 0.8, y = 0,
            text = "High-Capacity Outdoor Bike Parking Locations",  # Caption for parking_map
            showarrow = FALSE,
            xref = "paper", yref = "paper",
            font = list(size = 10, color = "black"),
            align = "center"
        )
    ),
    nrows = 1, margin = 0.05  # Arrange plots in one row
) %>%
    layout(
        title = list(
            text = "Toronto Neighbourhoods with More Bikes Have More Bike Thefts",
            x = 0.5  # Center the title
        )
    )

# Display the combined plot
combined_plot

```

When we compare the locations of reported bike thefts from 2023 and City of Toronto designated bike parking side-by-side, we see two very similar maps. Data about the number of cyclists and unattended bikes in each Toronto neighbourhoods is not readily available, so I have used the locations of City-owned high-capacity outdoor bike parking facilities to get an idea of which neighbourhoods have more cyclists, and therefore more unattended bikes.

This reveals some important insights about bike theft risk in different parts of Toronto:

-   

We saw that most bike thefts

Although we saw consistent trends in the breakdown of yearly bike thefts by premise type between 2014 and 2023 across the City of Toronto, different neighbourhoods have different high-theft premises. There is variation among the top 10. Most bike thefts in the Yonge-Bay corridor took place outside or by a commercial building; this makes sense because there are very few residential buildings in that area and many office buildings.

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-piechart
#| fig-cap: ["10 busy downtown neighbourhoods, out of Toronto's 158 neighbourhoods account for 41% of the bike thefts since 2014 due to high cyclist volume."]
# Ensure required libraries are loaded
library(tibble)
library(plotly)

# Data setup
graph_this_data <- tibble(
  category = c("10 Downtown Neighbourhoods", "The Other 148"),
  proportion = c(
    round((sum(top_10_neighbourhoods$num_thefts) / 
             n_distinct(bike_thefts_clean$EVENT_UNIQUE_ID)), 2),
    1 - round((sum(top_10_neighbourhoods$num_thefts) / 
                 n_distinct(bike_thefts_clean$EVENT_UNIQUE_ID)), 2)
  )
)

# Display the pie chart
graph_this_data <- graph_this_data %>%
  mutate(label = paste0(round(proportion * 100), "%"))

static_pie_chart <- ggplot(graph_this_data, aes(x = "", y = proportion, fill = category)) +
  geom_bar(stat = "identity", width = 1, color = "white") +
  coord_polar("y", start = 0) +
  # scale_fill_manual(values = c("#e7298a", "gray")) +
  scale_fill_manual(values = c("#EF553B", "#636EFA")) + # Custom colors
  geom_text(
    aes(label = label),
    position = position_stack(vjust = 0.5),
    size = 4,
    color = "white"
  ) +
  labs(
    title = "Where in Toronto are the most bikes stolen?",
    fill = ""
  ) +
  theme_void() + # Removes unnecessary chart elements
  theme(
    plot.title = element_text(hjust = 0.5, size = 12),
    legend.position = "bottom"
  )
static_pie_chart

```

Unsurprisingly, the ten neighbourhoods that have had the most bike thefts since 2014 are in downtown. But is your bike really at a higher risk of being stolen in the Yonge-Bay Corridor than any other Toronto neighbourhood? Not necessarily.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-hotspot
#| fig-cap: Bike theft premise trends in the 10 neighbourhoods with the most thefts are similar to the city overall. University is a notable outlier, with more than half of its bike thefts taking place on educational premises at the University of Toronto St. George Campus.
#| fig-cap-location: bottom

neighbourhood_counts <- bike_thefts_sf |>
  group_by(NEIGHBOURHOOD_158) |>
  summarise(num_thefts = n_distinct(OBJECTID),
            lat = mean(LAT_WGS84),
            long = mean(LONG_WGS84),
            .groups="drop") 

top_10_neighbourhoods2 <- top_10_neighbourhoods

top_10_neighbourhoods2$NEIGHBOURHOOD_158 <- gsub("\\s*\\(.*\\)", "", top_10_neighbourhoods2$NEIGHBOURHOOD_158)

neighbourhood_counts2 <- bike_thefts_sf %>%
  filter(NEIGHBOURHOOD_158 %in% top_10_neighbourhoods$NEIGHBOURHOOD_158) %>%
  group_by(NEIGHBOURHOOD_158, PREMISES_TYPE) %>%
  summarise(num_thefts = n_distinct(OBJECTID),  # Count distinct thefts
            lat = mean(LAT_WGS84),               # Mean latitude for the neighborhood
            long = mean(LONG_WGS84),             # Mean longitude for the neighborhood
            .groups = "drop") %>%
  group_by(NEIGHBOURHOOD_158) %>%
  mutate(total_thefts = sum(num_thefts),  # Total thefts per neighborhood
         percent_thefts = num_thefts / total_thefts * 100) %>%
  ungroup()  # Remove grouping after calculation

neighbourhood_counts2$NEIGHBOURHOOD_158 <- gsub("\\s*\\(.*\\)", "", neighbourhood_counts2$NEIGHBOURHOOD_158)

neighbourhood_counts2$PREMISES_TYPE <- factor(
  neighbourhood_counts2$PREMISES_TYPE,
  levels = c("Other", "Transit", "Educational", "Commercial", "Outside", "House", "Apartment" )# Adjust as needed
)

custom_colors <- c("#8dd3c7", "#ffffb3", "#bebada", "#fb8072", "#80b1d3", "#fdb462", "#b3de69")


ggplot(neighbourhood_counts2, aes(x = reorder(NEIGHBOURHOOD_158, num_thefts), y = num_thefts, fill = PREMISES_TYPE)) +
  geom_bar(stat = "identity", position="stack") +  # Bar plot where height of bars corresponds to num_thefts
  scale_fill_manual(name = "Premise Type", values = custom_colors) +
  coord_flip() +  # Flip the axes for better readability of neighborhood names
  labs(
    title = "Where do bike thefts happen downtown?",
    x = "Neighborhood",
    y = "",
    fill = "Premise Type"
  ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),  # Remove grid lines for a cleaner look
    axis.text.x = element_text(angle = 30, hjust = 1, size=8),  # Rotate x-axis labels if needed
    legend.position = "bottom"  # Place the legend at the bottom
  )
```

This section focuses on 2023 because the bike parking location data I obtained is from 2023. Bike parking locations in 2023 may not have existed back in 2014.

5 most affected neighborhoods and premises where bikes were stolen from in 2023

North York, near Yonge and Finch, saw a high concentration of bike thefts in 2023. This is particularly concerning because the bike parking facility data obtained from Open Data Toronto shows no City of Toronto bike parking facilities in that area or at that intersection. This indicates that cyclists who live in the neighbourhoods surrounding Yonge Street in North York may be at a higher risk of having their bikes stolen. If you are a cyclist who lives in North York near Yonge Street and you leave your bike outside at any time of day, it might be a good idea to invest in an effective bike lock.

Your bike is not more likely to be stolen downtown

```{r}
#| echo: false
#| warning: false
#| message: false
#| include: false
library(leaflet)
leaflet() %>%
    # addProviderTiles("OpenStreetMap") %>%
    addProviderTiles('CartoDB.Positron') %>%
    addPolygons(data = toronto_neighbourhoods_shp, weight = 1, color = "gray", fillOpacity = 0.2) %>%
    addCircleMarkers(data = bike_sf, radius = 0.5, color = "#980043", fillColor = "#980043", popup = ~NEIGHBOURHOOD_140) %>%
    addLegend(position = "bottomright", colors = "#980043", labels = "Bike Theft")

```

```{r, include=FALSE, echo=FALSE, warning=FALSE, message=FALSE}
library(leaflet)
library(dplyr)

# Convert the sf object to a data.frame
bike_parking_df <- bike_parking_sf %>%
  st_transform(crs = 4326) %>%  # Ensure it's in the correct coordinate reference system (WGS 84)
  st_coordinates() %>%  # Extract coordinates
  as.data.frame() %>%
  bind_cols(bike_parking_sf)  # Combine the coordinates with the original data

# Plot the map with bike parking location dots
leaflet(toronto_neighbourhoods_shp) %>%
    addProviderTiles('CartoDB.Positron') %>%
    # Add the neighborhood polygons
    addPolygons(weight = 1.2, fillOpacity = 0.1, color = "gray") %>%
    # Add the bike parking location dots
    addCircleMarkers(
      data = bike_parking_df,  # Use the data frame with coordinates
      lng = ~X, lat = ~Y,  # Specify longitude (X) and latitude (Y)
      radius = sqrt(bike_parking_df$BICYCLE_CAPACITY),
      color = "steelblue",  # Color of the dot
      fillColor = "steelblue",  # Fill color
      fillOpacity = 0.7,  # Opacity of the fill,
      stroke = FALSE  # No border for the dots
      # Replace with the correct attribute
    ) %>%
    addLegend(position = "bottomright", colors = "steelblue", labels = "City of Toronto Bike Parking Facility")

```

```{r, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
neighbourhood_recovery <- recovered_bikes_top10
neighborhood_recovery_2023 <- recovered_bikes_top10 |>
  filter(REPORT_YEAR == 2023)  # Filter data for the year 2023

# Remove parentheses and numbers from neighborhood names
neighborhood_recovery_2023$NEIGHBOURHOOD_158 <- gsub("\\s*\\(.*\\)", "", neighborhood_recovery_2023$NEIGHBOURHOOD_158)

# Create the bar graph
ggplot(neighborhood_recovery_2023, aes(x = reorder(NEIGHBOURHOOD_158, num_thefts), y = pct_recovered)) +
  geom_bar(stat = "identity", fill = "#80b1d3", color = "black") +
  scale_y_continuous(labels = scales::number_format(accuracy = 0.1, suffix = "%")) +  # Format y-axis with % symbol
  coord_flip() +  # Flip the axes for better readability
  labs(
    title = "Percentage of Bikes Recovered by Neighborhood (2023)",
    x = "Neighborhood",
    y = "Percentage Recovered"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16),
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size=10)
  )
```

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-downtowngone
#| fig-cap: Only 0.73% of stolen bikes were recovered in 2023, but in 8 of the 10 neighbourhoods with the most bike thefts since 2014, no stolen bikes were recovered in 2023. While the risk of bike theft is not higher here, the chances of recovering a stolen bike are lower.

neighborhood_recovery_2023_long <- neighborhood_recovery_2023 %>%
  select(NEIGHBOURHOOD_158, num_recovered, num_not_recovered) %>%
  pivot_longer(cols = c(num_recovered, num_not_recovered), 
               names_to = "status", 
               values_to = "count") %>%
  mutate(
    status = recode(status, 
                    num_recovered = "Recovered", 
                    num_not_recovered = "Not Recovered")
  ) %>%
  group_by(NEIGHBOURHOOD_158) %>%
  mutate(percentage = count / sum(count)) %>%  # Calculate percentages
  ungroup()

neighborhood_recovery_2023_long$status <- factor(neighborhood_recovery_2023_long$status, levels = c("Recovered", "Not Recovered"))


# Create the stacked bar graph
plot <- ggplot(neighborhood_recovery_2023_long, aes(x = reorder(NEIGHBOURHOOD_158, count), y = percentage, fill = status)) +
  geom_bar(stat = "identity", color = "black", size = 0.25,
           aes(text = scales::percent(percentage))) +  # Set thinner outline
  scale_fill_manual(
    name = "Status",
    values = c("Recovered" = "#80b1d3", "Not Recovered" = "#fb8072")
  ) +
  coord_flip() +  # Flip axes for better readability
  labs(
    title = "If Your Bike Gets Stolen Downtown, You're Really Not Getting it Back",
    x = "Neighborhood",
    y = "Percentage"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 12),
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 10)
  ) +
  scale_y_continuous(labels = scales::percent_format())

interactive_plot <- ggplotly(plot, tooltip = "text")
interactive_plot

```

While living in one of these neighbourhoods may not put you at a higher risk of having your bike stolen, the already slim chance of finding a stolen bike is virtually non-existent.

# Data-Driven Tips for Toronto Cyclists

Even though bike thefts are on the rise, over 99% of bikes stolen since 2021 were never found. Unfortunately, once a bike gets stolen, there is not much you can do. Prevention is key, so here are some things every Toronto cyclist should consider:

-   **Your bike is probably no more likely to get stolen, but even less likely to be recovered if it is stolen downtown** than in any other part of Toronto. 0.73% of stolen bikes across Toronto were recovered in 2023, but not a single bike stolen in the busy Yonge-Bay corridor has been recovered since 2021.

-   **Bike thieves love to target apartments and houses**. If you live in an apartment, consider keeping your bike inside a your unit or your building's bike storage facility. I know that this is not an option for every cyclist in an apartment, so if you really have to leave your bike outside consider using two high-quality bike locks. If you live in a house, consider storing your bike inside a shed, garage, or your house instead of your front porch or anywhere else outside.

-   **Double-lock your bike,** especially if you leave it on an outdoor rack or pole, or anywhere on the University of Toronto's St. George Campus. One limitation of this suggestion is that information about the percentage of stolen bikes that were locked and what type of lock was used is not available, but it doesn't hurt to take extra precautions.

-   There is reason to believe that **bike thieves love the** **University of Toronto St. George Campus.** Nearly half the bike thefts reported in the University neighbourhood since 2014 took place at an educational premise. If you're a student rushing between classes, securing your bike is absolutely worth it. Students and faculty at the University of Toronto can access a Bike Enclosure Room on campus, so this might be something to consider as well.

-   There is reason to believe that **risk of bike theft may be elevated in North York.** The highest concentration of high-capacity bike parking facilities and bike thefts is in 10 busy downtown neighbourhoods. However, there is a cluster of 2023 bike thefts in North York, specifically the Yonge-Doris, East Willowdale, Willowdale West, and Newtonbrook West neighbourhoods. These neighbourhoods have zero high-capacity bike parking facilities, so the number of thefts there might be disproportionately high. One limitation of this is that locations of high-capacity bike parking faciities aren't a perfect way to measure cyclist volume. Locations of high-capacity bike parking were used to get an idea of high cyclist volume areas because the City of Toronto may be more inclined to build these facilities where there is demand for them.
